\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-30 syntax.tex
resolve comments

A term $\goodTerm{\Gamma}{t}{A}$ can be understood as one term $t$ of type $A$, with a bunch of holes (of type $\Gamma$) inside the term $t$. A substitution $\goodSub{\Gamma}{\gamma}{\Delta}$ can be similarly understood in the way that $\gamma$ will output a \textbf{series} of (dependently-typed related) terms of types $\Delta$, parametrized by $\Gamma$ (i.e. we can consider there are holes in $\gamma$ of type $\Gamma$). Thus (1) \ruleref{Sub Extend} looks like concatenating one series of terms with a new term (the weird looking type of $\goodTerm{\Gamma}{t}{A[\sigma]}$ is because we need $A$ to appear in the context of $\Delta$ so that we can use \ruleref{Context Extension} to create new context); (2) we have \ruleref{Proj Subst-1} to extract a subpart of a given substitution; (3) \ruleref{Proj-Ext} explains why substitution really looks like a (deeply-nested) pair.

Since the $\Gamma$ before the $\vdash$ is always indicating the types of ``holes'', this explains the \ruleref{Term Subst} and \ruleref{Type Subst}: given a term $\goodTerm{\Delta}{t}{A}$ and a substitution $\goodSub{\Gamma}{\gamma}{\Delta}$, we will have a new term $\goodTerm{\Gamma}{t[\gamma]}{A[\gamma]}$ after the substitution---this substitution can be understood as ``filling all the holes''. Once all the holes are filled with a given substitution, the new holes all come from the subsitution itself. Thus consecutive subsitution can be composed as \ruleref{Sub Comp}, because all the holes are always from the last substitutions.

Note that, for simplicity, we will (a lot of time) writing $\pi_1$ as a \textbf{shorthand} of $\goodSub{\Gamma,A}{\pi_1~"id"}{\Gamma}$, and similarly another shorthand for $\goodTerm{\Gamma, A}{\pi_2 := \pi_2~"id"}{A[\pi_1]}$.

Based upon the basic MLTT,
we will have four new kinds of judgements to model our linkages. Among them, sealing judgement $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is the
most special one because it is actually just a shorthand of term
judgement $\goodTerm{\Gamma , \cC \sigma}{\_}{A[\pi_1]}$.\YZ{You haven't introduced the notation A[pi_1] yet at this point. And what is A[pi_1]?}\EDJreply{added for the intro. It is a shorthand for pi_1(id)}

@@

We will have dependent function types
and dependent pair types. Dependent pair type is used to model Coq
module, which is useful when we need to model the ``compilation'' from
linkages to modules. We also have $\beta,\eta$-rule for both dependent
function type and dependent pair type.
The substitution law for these two
require \textit{de bruijn lifting}: for a given substitution
$\goodTerm{\Gamma}{\sigma}{\Delta}$ and a type
$\goodType{\Delta}{A}{}$, we define its de Bruijn lifting
$\goodTerm{\Gamma,A[\sigma]}{\sigma^\uparrow := (\sigma \circ \pi_1, \pi_2)}{\Delta, A}$ . Intuitively speaking, de Bruijn lifting is to ``fill all the holes except the last one''.\YZ{What is A[sigma] and what is de Bruijn lifting? Seems like if the reader is not familiar with the substitution calculus, then they will get lost starting here. Broadly speaking, a gentle intro to the intrinsic setting would help the reader a great deal.}\EDJreply{A quick intuitive explanation of substitution-calculus is given above. Please check. }

Because in dependent type, function now can return type, of which the type is ``type of type'', \textit{the universe} $\cU$. We follow the style of \textit{type universe à la Tarski}~\cite{hofmann1997syntax} and
distinguish types from their \textit{codes or names}, and thus we use
$El\ T$ as the type given \textit{type name} $T : \cU$. In this case, (1) $\cU$ is the type including the \textit{codes or names} of types , and (2) $T$ is
not allowed to locate at type position (i.e. at the right-most position of a term judgement) but $El \ T$ is. "c" is the ``inverse'' of $El$ that getting \underline{"c"}ode of a type, with the following two equations $"c" (El~T) \equiv T$ and $El ("c"~T) \equiv T$. Recall that we are actually using predicative universe and thus we actually have different level of universes $\cU_j$. Here level omitted for conciseness.

@@

$$
\quad 
\Rule[name=Func/DPair Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}
\quad \goodType{\Delta}{A}{} 
\quad \goodType{\Delta, A}{B}{}
}
{
  \goodType{\Gamma}{(\Pi A B)[\gamma] \equiv \Pi A[\gamma] B[\gamma^\uparrow] }{j \cup k}
  \quad 
  \goodType{\Gamma}{(\Sigma A B)[\gamma] \equiv \Sigma A[\gamma] B[\gamma^\uparrow] }{j \cup k}
}
$$
%\YZ{What is the requirement on Delta?}
%\EDJreply{ Delta here is of type Con and thus Delta here should be a well-typed Context. This is another issue I want to talk about, in Agda these kinds of stuff is inferred (so the reader can infer them as well). For example, we know delta is at the place of context, so it is a context, but we don't need to specify that it is a ``well-typed context'' because every context is well-typed in this intrinsic-typing. What's more, I think adding this extra judgement here just cause significant code bloat (like what I complained to you before I know how to use implcit variable in agda. For example, TYPE SUBST, there will be two more judgements (with no more information there) ) (Another reason I prefer fake agda formulation) }\YZreply{I was thinking about 'Delta |- A'. I suppose it is also implicitly enforced via intrinsically typed syntax. Seems to me it is worth stating, however, or otherwise the rule appears to say that Delta can be any context as long as Gamma |- gamma : Delta.}\EDJreply{You are absolutely right! I think Delta |- A, even though inferrable, mentioning it is good, also some information are B is also ok, please check, resolve if ok, thanks!}
$$
\Rule[name=Type Subst]
{\goodType{\Delta}{T}{j} 
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodType{\Gamma}{T[\gamma]}{j}}
\quad
\Rule[name=Base Type Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}}
{\goodType{\Gamma}{\cU[\gamma] \equiv \cU }{j + 1} \quad
  \goodType{\Gamma}{\cB[\gamma] \equiv \cB}{0} \quad 
  \goodType{\Gamma}{\bot[\gamma] \equiv \bot}{0}
}
$$
\judgebox{ \goodTerm{\Gamma}{t}{T} }
$$
\Rule[]
{\goodType{\Gamma}{T}{j}}
{\goodTerm{\Gamma}{"c" \ T}{\cU}\YZ{what is c? Why is this rule called Univ-1?Didn't you write that you omit universe levels?}\EDJreply{I think c stands for code. I am just following convention. Now I remove their name because that is a bad name. About the level, I just kind of afraid they will think there is one universe. Now I've removed all the indices}
}\quad
\Rule[]
{\goodTerm{\Gamma}{T}{\cU}}
{\goodType{\Gamma}{El \ T}{j}\YZ{'El T' isn't explained only until it makes a second appearance in Ind-Term.}\EDJreply{I have made El and universe explanation earlier, please check.}}
\quad
\Rule
{}
{\goodTerm{\Gamma}{"tt", "ff"}{\cB}}
\quad 
\Rule[name=Term Subst]
{\goodTerm{\Delta}{t}{T}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodTerm{\Gamma}{t[\gamma]}{T[\gamma]}}
$$
$$
\Rule[]
{\goodTerm{\Gamma, A}{t}{B}}
{\goodTerm{\Gamma}{\lambda t}{\Pi A B}}
\quad 
\Rule[]
{\goodTerm{\Gamma}{u}{A} 
\quad \goodTerm{\Gamma}{v}{B[(id, u)]}}
{\goodTerm{\Gamma}{(u,v)}{\Sigma A B}}
\quad 
\Rule[]
{\goodTerm{\Gamma}{t}{\Sigma A B}}
{\goodTerm{\Gamma}{"pjl" \ t}{A}
\quad \goodTerm{\Gamma}{"pjr" \  t}{B[(id, "pjl" \  t)]}
}
$$
$$
\Rule
{}
{\goodTerm{\Gamma}{(\lambda t)[\gamma] \equiv \lambda (t[\gamma^\uparrow])}{\Pi A B}
\quad \goodTerm{\Gamma}{(u,v)[\gamma] \equiv (u[\gamma],v[\gamma])}{\Sigma A B}
\quad \goodType{\Gamma}{El \ (T[\gamma]) \equiv (El \ T) [\gamma]}{}
}
$$

@@

We have shown case some exemplar rules for all four standard type
judgments, and now we focus on the newly introduced facility.

@@

\caption{Seal Judgement, and its Agda Representation \\ (or What do we mean when we say one judgement is a shorthand of the other)}

@@

As we said earlier, $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is actually just
a shorthand of term judgement $\goodTerm{\Gamma , \cC
\sigma}{\_}{A[\pi_1]}$,\YZ{Not quite. Previously you used cL I think.}\EDJreply{Sorry. It is fixed. It should be $\cC$ in the context.}
a special form of \textit{local morphism}~\cite{abbott2003category}.
The substitution law should be understood as a lemma, and it is easily
derivable using $\gamma^\uparrow$.

>>>>

A term $\goodTerm{\Gamma}{t}{A}$ can be understood as one term $t$ of
type $A$, with a bunch of holes (of type $\Gamma$) inside the term $t$.
A substitution $\goodSub{\Gamma}{\gamma}{\Delta}$ can be similarly
understood in the way that $\gamma$ will output a \textbf{series} of
(dependently-typed related) terms of types $\Delta$, parametrized by
$\Gamma$ (i.e., we can consider there are holes of type $\Gamma$ in $\gamma$).
Thus (1) \ruleref{Sub Extend} looks like concatenating one
series of terms with a new term (the weird looking type of
$\goodTerm{\Gamma}{t}{A[\sigma]}$ is because we need $A$ to be typed in
the context $\Delta$ so that we can use \ruleref{Context Extension} to
create new context); (2) we have \ruleref{Proj Subst-1} to extract a
subpart of a given substitution; (3) \ruleref{Proj-Ext} explains why
substitution really looks like a (deeply-nested) pair.

Since the $\Gamma$ before the $\vdash$ is always indicating the types of
``holes'', this explains the \ruleref{Term Subst} and \ruleref{Type
Subst}: given a term $\goodTerm{\Delta}{t}{A}$ and a substitution
$\goodSub{\Gamma}{\gamma}{\Delta}$, we will have a new term
$\goodTerm{\Gamma}{t[\gamma]}{A[\gamma]}$ after the substitution---this
substitution can be understood as ``filling all the holes''. Once all
the holes are filled with a given substitution, the new holes all come
from the substitution itself.
Consecutive substitutions can be composed using \ruleref{Sub Comp}, because
all the holes are always from the last substitutions.

Note that, for simplicity, we will often write $\pi_1$ as a
shorthand for ${\pi_1~"id"}$, typed as
$\goodSub{\Gamma,A}{\pi_1~"id"}{\Gamma}$, and $\pi_2$ as a shorthand
for $\pi_2~"id"$, typed as $\goodTerm{\Gamma, A}{\pi_2~"id"}{A[\pi_1]}$.

The core language builds on MLTT and adds four new judgement forms.
Among them, sealing judgement $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is just a
shorthand for term judgement
$\goodTerm{\Gamma , \cC \sigma}{\_}{A[\pi_1]}$.

@@

We will have dependent function types
and dependent pair types. Dependent pair type is used to model Coq
module, which is useful when we need to model the ``compilation'' from
linkages to modules. We also have $\beta,\eta$-rule for both dependent
function type and dependent pair type.
The substitution law for these two
require \textit{de Bruijn lifting}: for a given substitution
$\goodTerm{\Gamma}{\sigma}{\Delta}$ and a type
$\goodType{\Delta}{A}{}$, we define its de Bruijn lifting
$\goodTerm{\Gamma,A[\sigma]}{\sigma^\uparrow := (\sigma \circ \pi_1, \pi_2)}{\Delta, A}$ . Intuitively speaking, de Bruijn lifting is to ``fill all the holes except the last one''.

Because in dependent type, function now can return type, of which the type is ``type of type'', \textit{the universe} $\cU$. We follow the style of \textit{type universe à la Tarski}~\cite{hofmann1997syntax} and
distinguish types from their \textit{codes or names}, and thus we use
$El\ T$ as the type given \textit{type name} $T : \cU$.
In this case, (1) $\cU$ is the type including the \textit{codes or
names} of types, and (2) $T$ is not allowed to locate at type position
(i.e., at the rightmost position of a term judgement) but $El \ T$ is.
"c" is the ``inverse'' of $El$, obtaining  the \underline{"c"}ode of a
type, with the following two equations $"c" (El~T) \equiv T$ and $El
("c"~T) \equiv T$.
Recall that we are actually using predicative universe, so we actually
have different levels of universes $\cU_j$.
We omit universe levels for conciseness.

@@

$$
\quad 
\Rule[name=Func/DPair Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}
\quad \goodType{\Delta}{A}{} 
\quad \goodType{\Delta, A}{B}{}
}
{
  \goodType{\Gamma}{(\Pi A B)[\gamma] \equiv \Pi A[\gamma] B[\gamma^\uparrow] }{j \cup k}
  \quad 
  \goodType{\Gamma}{(\Sigma A B)[\gamma] \equiv \Sigma A[\gamma] B[\gamma^\uparrow] }{j \cup k}
}
$$
%\YZ{What is the requirement on Delta?}
%\EDJreply{ Delta here is of type Con and thus Delta here should be a well-typed Context. This is another issue I want to talk about, in Agda these kinds of stuff is inferred (so the reader can infer them as well). For example, we know delta is at the place of context, so it is a context, but we don't need to specify that it is a ``well-typed context'' because every context is well-typed in this intrinsic-typing. What's more, I think adding this extra judgement here just cause significant code bloat (like what I complained to you before I know how to use implcit variable in agda. For example, TYPE SUBST, there will be two more judgements (with no more information there) ) (Another reason I prefer fake agda formulation) }\YZreply{I was thinking about 'Delta |- A'. I suppose it is also implicitly enforced via intrinsically typed syntax. Seems to me it is worth stating, however, or otherwise the rule appears to say that Delta can be any context as long as Gamma |- gamma : Delta.}\EDJreply{You are absolutely right! I think Delta |- A, even though inferrable, mentioning it is good, also some information are B is also ok, please check, resolve if ok, thanks!}
$$
\Rule[name=Type Subst]
{\goodType{\Delta}{T}{j} 
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodType{\Gamma}{T[\gamma]}{j}}
\quad
\Rule[name=Base Type Subst]
{\goodSub{\Gamma}{\gamma}{\Delta}}
{\goodType{\Gamma}{\cU[\gamma] \equiv \cU }{j + 1} \quad
  \goodType{\Gamma}{\cB[\gamma] \equiv \cB}{0} \quad 
  \goodType{\Gamma}{\bot[\gamma] \equiv \bot}{0}
}
$$
\judgebox{ \goodTerm{\Gamma}{t}{T} }
$$
\Rule[]
{\goodType{\Gamma}{T}{j}}
{\goodTerm{\Gamma}{"c" \ T}{\cU}
}\quad
\Rule[]
{\goodTerm{\Gamma}{T}{\cU}}
{\goodType{\Gamma}{El \ T}{j}}
\quad
\Rule
{}
{\goodTerm{\Gamma}{"tt", "ff"}{\cB}}
\quad 
\Rule[name=Term Subst]
{\goodTerm{\Delta}{t}{T}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodTerm{\Gamma}{t[\gamma]}{T[\gamma]}}
$$
$$
\Rule[]
{\goodTerm{\Gamma, A}{t}{B}}
{\goodTerm{\Gamma}{\lambda t}{\Pi A B}}
\quad 
\Rule[]
{\goodTerm{\Gamma}{u}{A} 
\quad \goodTerm{\Gamma}{v}{B[(id, u)]}}
{\goodTerm{\Gamma}{(u,v)}{\Sigma A B}}
\quad 
\Rule[]
{\goodTerm{\Gamma}{t}{\Sigma A B}}
{\goodTerm{\Gamma}{"pjl" \ t}{A}
\quad \goodTerm{\Gamma}{"pjr" \  t}{B[(id, "pjl" \  t)]}
}
$$
$$
\Rule
{}
{\goodTerm{\Gamma}{(\lambda t)[\gamma] \equiv \lambda (t[\gamma^\uparrow])}{\Pi A B}
\quad \goodTerm{\Gamma}{(u,v)[\gamma] \equiv (u[\gamma],v[\gamma])}{\Sigma A B}
\quad \goodType{\Gamma}{El \ (T[\gamma]) \equiv (El \ T) [\gamma]}{}
}
$$

@@

We have shown some exemplar rules for all four standard types of
judgments. Now we focus on the newly introduced facility.

@@

\caption{Seal Judgement, and its Agda Representation \\ (or What do we mean when we say one judgement is a shorthand for the other)}

@@

As we said earlier, $\goodSeal{\Gamma}{\_}{\sigma}{A}$ is actually just
a shorthand for term judgement $\goodTerm{\Gamma , \cC
\sigma}{\_}{A[\pi_1]}$,\YZ{Not quite. Previously you used cL I think.}\EDJreply{Sorry. It is fixed. It should be $\cC$ in the context.}
a special form of \textit{local morphism}~\cite{abbott2003category}.
The substitution law should be understood as a lemma, and it is easily
derivable using $\gamma^\uparrow$.
